Reachability analysis for continuous one-counter automata

### 2021-08-17 Documentation/paper.tex
lots of fixes

@@

\title{Reachability analysis for continuous one-counter automata}
\author{Lars Van Roy\\
\textit{dept. of Mathematics and Computer Science} \\
\textit{University of Antwerp}\\
lars.vanroy@student.uantwerpen.be}

\begin{document}
\maketitle{}

\begin{abstract}
\noindent
Reachability is a known problem that can heavily affect the efficiency of code. It has been proven that reachability is decidable for continuous one counter timed automata which we will utilise in our approach. The proposed approach will provide a means to convert c code to one counter timed automata followed by an algorithm which is capable of analysing the reachability of each of the nodes within the generated automata. This is an extension to existing research in which we already provided a means to perform the conversion from code to automata. Within this paper, this approach will be repeated here, as the approach is specifically intended to be combined with this predecessor. Finally, we will apply the combination of approaches on an existing code base, as to determine the usefulness of the approach.

@@

\section{Introduction}
A well known problem in writing code is the reachability of specific lines in said written code. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code \cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired, but is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached. to remedy this, a proper reachability prevention approach is needed.

To analyse the reachability, we will make use of the fact that code is quit easily convertible to automata. All statements within the code that are related to counter updates can directly be converted to operations within an automaton. Each line of code can represented by one or more nodes and a traversal of an automaton can be directly linked to the traversal of the code itself.

The continuous one counter automata are especially interesting as it has been proven that reachability is decidable for this type of automata \cite{succintandparametriconecounter, danialandjoel, blondin2021continuous}. In this paper we will provide a practical implementation for the theoretical approach that was provided by Tim Leys et al.\cite{blondin2021continuous} which can be used to prove reachability of nodes.

Adding the restriction of only allowing a single parametric clock has a big effect on the types of code we can analyse. However, this is unavoidable, since there is no proof for two counter machines. Three counter machines and up have been proven to be undecidable \cite{FEARNLEY201526}.

In this paper we will provide the overview of an approach which will convert c code to a one counter automaton. Using the result of this first step, we will then provide a means to analyse the reachability of any of the nodes of said automaton. The following steps are taken without our approach to prove the existence of dead code.

\begin{enumerate}
	\item Convert c code into a parse tree, using ANTLR
	\item Perform context free reductions on this parse tree
	\item Perform context sensitive reductions on this parse tree
	\item Validate the resulting parse tree, to be conforming to the code constraints
	\item Generate the counter automaton
	\item Analyse which nodes are reachable
	\item Link potential unreachable statements back to the lines within the original code base
\end{enumerate}

@@

Other then the described code constraint, functions must also have a boolean return type, and integer parameter types (if any) to be converted into a counter automaton.

\section{Motivation}
Popular IDE's do already implement reachability analysis up to a certain degree. However, the dead code elimination they perform is shallow and limited. The code example given below was executed using Clion, one of

\begin{lstlisting}[style=CStyle]
bool is_mul(int a, int b) {
  int temp = b;
  for(; temp > 0;){
    if(temp == 0){
      break;
    }
    temp -= a;
  }
  return temp == 0;
}

@@

\noindent
the more commonly used C IDE's. While it is known that Clion can find dead code that directly results from code, it was not able to resolve this case, where the condition on line 4 can never evaluate to true. This is because the for statement on line 3 will always prevent line 4 from evaluating to true and yet Clion did not detect this.

@@

The impact of removing this line can be significant if considering cases where bigger numbers are passed. In those cases the loop on line 3 will be executed several times, making the condition on line 4 being quit impactful on the total evaluation time of this function. This clearly shows that there is value in enhanced dead code elimination, compared to the current standards.

\section{Preliminaries}
This paper will make use of counter automata and more specifically one counter automata. These automata are derived from the general definition of automata and signify an extension by adding the notion of counters.

@@

At all times the automaton will be in a configuration $(q, c)$ where $q \in Q$ and $c$ the counter valuation where $(q_0, 0)$ is the initial configuration. The counter valuation $c \in \mathbb{N}$ represent one single value for which the counter is evaluated. In our analysis the goal will be to find the range of possible counter values that the counter can have in each state $q \in Q$.

The $\tau$ parameter represents the mapping from all states to the interval of possible counter values that can exist in that state. For example, if a node $q$ has an associated condition stating $\leq 4$ the $\tau$ function would have entry for node $q$ equal to $(-\infty, 4]$. The set of supported constraints within our evaluation are visualised by the $cond$ set below with $\epsilon$ representing the empty condition. Note that this is not binding for the code that we can analyse, several conditions might be convertible to constructs using only supported statements.
\[
cond = \{\leq c, =c, \geq c, \leq p, =p, \leq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

Parametric counter automata, as compared to regular counter automata, allow counters to be modified by parameters as well as constants. As the counters need to be integers, we will only allow integer type parameters. We will consider the following set of allowed operations where $P$ is the set of parameters, the $\epsilon$ symbol will be used to represent an empty label.
\[
op = \{+c, -c, +p, -p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

The final state is omitted from this definition as we are not interested in this. In general the reachability of every single line of code will be under scope and not a single one of the lines. Furthermore, making the assumption that the final state is the only state of which we want to analyse reachability creates a significant reduction in the use cases of this approach, and was therefore discarded.

\subsection{COCA reachability}
In this section we give an overview of how we approach the concept of reachability. The provided concept was proven by Michael Blonding et al. \cite{blondin2021continuous} and will be the basis for the remainder of the approach.

Throughout the evaluation we will be tracking an interval $I$ which will have $\vert Q \vert$ entries each corresponding to the counter values that can exist within their respective node.

The concept of reachability will be associated with whether or not a certain node has a non empty interval at the end of the evaluation. A non empty interval automatically implies that the node must be reachable cause if a node can have a counter value, it recursively implies that there must be a path $\pi$ in which all preceding nodes also have non empty intervals.

For this evaluation we will allow transitions to be partially taken. This means that every transition can be scaled by an unknown factor $\alpha \in (0,1]$. This implies that given that we have a preceding interval $[0, 3]$ and an operation $+2$ our resulting interval will not be $[2, 5]$ but rather $(0, 5]$ as with $\alpha \approx 0$ we would remain at zero and with $\alpha = 2$ we would be able to go up to 5. The zero is not included within the resulting interval as we do not allow our $\alpha$ to actually be zero.

The partial aspect of transitions leads to potential false positives. In the example of $[0, 3] + 2 \rightarrow (0, 5]$ we might later on assume that counter values can be 0, which is not necessarily the case. We do however guarantee that in case we decide that a node is not reachable, this does hold. This implies that we may not be able to prove the reachability of every line, but for every line where we can prove it we know for certain that it is correct.

\section{C code to automata conversion}

The approach given within this paper consists out of two major phases. The first phase will consist of the conversion from c code to one counter automata. The second phase will use the result from the first phase, and perform a reachability analysis on these generated automata.

\begin{figure}[t]
	\makebox[\textwidth][c]{
		\begin{tikzpicture}[auto, >=latex, font=\footnotesize\scshape]
		\tikzstyle{center} = [draw=black, rectangle, text width=5.8em]

		\node[center]	at (0, 0) 		(q0) 	{Generate tree};
		\node[center]	at (3.5, 0)		(q1)	{Context-free \\optimization};
		\node[center]	at (7, 0)		(q2)	{Context-sensitive \\optimization};
		\node[center]	at (10.5, 0)	(q3)	{Validate conditions};
		\node[center]	at (14, 0)		(q4)	{Generate \\automaton};
		\node[center]	at (14, -2)		(q5)	{Discard \\tree};

@@

		\end{tikzpicture}
	}
\end{figure}

The first phase starts with an initialization, which consists of the generation of a parse tree. This will be done using the default c grammar, with the extension of the boolean type as this is not within the c language by default.

Using the resulting parse tree we will perform a simple context-free optimization where we will remove all needless nodes. This optimization will help us in the third step, where we will do a more in depth optimization cycle. This third cycle will perform optimizations based upon the context in which the nodes appear, and will try to simplify the syntax tree as much as possible. The resulting tree will contain no more potentially ambiguous nodes and all remaining nodes will be those relevant in the remainder of the evaluation.

The fourth step will be an analysis on whether or not the resulting abstract syntax tree conforms to all earlier specified requirements and, if this is the case, we will continue with the fifth step, which will be the generation of the counter automaton. For a function to conform it must have a boolean return value, the arguments (if any) must be of type integer and must never be mutated within the function body, there can only be one counter and the counter can only be mutated using the operations part of the $op$ set displayed below.
\[
op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

@@

\subsection{Initialization}
The program will start of by running the generator ANTLR compilation unit on the given c code. In case there are any issues regarding simple compilation rules, the program will exit with an error, indicating the issues in the code. For the remainder of the execution, the code is assumed to be correct, and there will be no regard for potential issues within the code.

@@

However, the creation of nodes is not the same in all situations, in some cases, a specific kind of node will be created, and in other cases a specific kind of node will not be created, depending on the rule used for the current node (which we can determine from the context of the current node). In order to know whether or not we need to pop a node from the node stack, we will allow the code to look at the top node in the stack, and evaluate the type of set node. No other context related evaluations will be allowed.\\
\\
\noindent
\textbf{Context-free analysis assignmentExpression example}\\
To look at a concrete example, consider the following rule in the grammar:
\begin{align*}
	&assignmentExpression\\
	&\text{:}\indent \text{conditionalExpression}\\
	&|\indent \text{unaryExpression assignmentOperator assignmentExpression}\\
	&|\indent \text{Digitsequence}; \\
\end{align*}
The rule for conditionalExpression needs to be there, as there is a chain of expressions, and some expressions need to be higher in the order than others, therefore all expressions with a higher priority need to be evaluated before the lower priority expressions. This is done by adding a rule to a lower priority expression in each expression class. It is however undesirable for these nodes to be added. Without regard for context in other nodes, we can simply look at the current node to see whether or not the conditionalExpression line is used (which ANTLR generated context allows us to do).

@@

Important to note is that no folding/substitution will occur within loops, as these operations occur a variable number of times, and since there is no way of determining the exact number of times without evaluating the expression, this is considered too complex and often impossible, as counters may depend on parameters, which need to be chosen at evaluation time.\\
\\
\noindent
\textbf{Context-sensitive reduction example: constant propagation}\\
Another example of a needed reduction, is constant propagation and folding. It is possible that a code segment would initially be rejected, due to use of unwanted variables, while in essence, these variables are nothing but variable representations of results of constant expressions. Consider the code snippet below. It would initially be rejected, because the counter variable is initialized with a variable that is neither a constant nor a parameter.

\begin{lstlisting}[style=CStyle]
bool divide(int p, int n){
	int variable = 5;
	variable += 20;
	int counter = variable;
	while(counter > 0){
		counter -= n;
	}
	counter += 1;
	counter -= 1;
	if(counter == 0){
		return true;
	}
	return false;
}
\end{lstlisting}

However, due to constant propagation and folding, we can simplify the two statements on line 2 and line 3, so that they become.

\begin{lstlisting}[style=CStyle]
int variable = 25;
\end{lstlisting}

Now that the variable is just a constant, we can substitute this variable in the declaration of counter, so that we now have the following.
\begin{lstlisting}[style=CStyle]
int variable = 25;
int counter = 25;
\end{lstlisting}

The new code segment is acceptable, but we will not stop here, as the first line of the code is useless. It has no further benefit for the execution and will therefore be dropped. The final counter assignment will not be folded.

>>>>

\usepackage[hidelinks]{hyperref}

\usepackage{datetime}

@@

\title{Reachability analysis for continuous one-counter automata}
\author{Lars Van Roy\\
\textit{dept. of Mathematics and Computer Science} \\
\textit{University of Antwerp}\\
lars.vanroy@student.uantwerpen.be\\
\vspace*{10em}
Supervisor}

\newdateformat{monthyeardate}{\monthname[\THEMONTH], \THEYEAR}

\begin{document}

\begin{titlepage}
	\begin{center}
		\vspace*{1cm}

		{\Large\textbf{Reachability analysis for continuous one-counter automata}}

		\vspace{1.5cm}

		\includegraphics[width=0.3\textwidth]{ua.jpg}

		\vspace{1.5cm}

		\textbf{Lars Van Roy}

		\vspace{0.5cm}

		\textit{dept. of Mathematics and Computer Science} \\
		\textit{University of Antwerp}

		\vspace{3cm}
		Supervised by \\
		Guillermo Alberto Perez \\
		Tim Leys

		\vspace{2cm}
		\monthyeardate\today
	\end{center}
\end{titlepage}

\begin{abstract}
\noindent
Reachability is an important metric and lack thereof is a known problem that can heavily affect the efficiency of code. It has been proven that reachability is decidable for continuous one counter automata which we will use in our approach. The proposed approach will provide a means to convert C code to continuous one counter automata followed by an algorithm which is capable of analysing the reachability of each of the nodes within the generated automata. This is an extension to existing research in which we already provided a means to perform the conversion from code to automata.

The newly added approach to compute reachability will make use of reachability intervals, which will be used to represent the set of values which a conditional variable can be equal to at any point throughout the evaluation of the automaton. There will be two variations, one for parametric counter automata and for non-parametric counter automata.

Finally, we will apply the combination of approaches on an existing project called xrdp, as to determine the usefulness of the approach. This application resulted in a total of 33 functions being identified as analysable of which all were correctly identified as fully reachable. To ensure that non reachable cases can also be handled further manual evaluation was performed which resulted in correct identification for all considered cases.

@@

\cleardoublepage
\phantomsection
\addcontentsline{toc}{section}{\listfigurename}
\listoffigures

\cleardoublepage <*-+/

\phantomsection
\addcontentsline{toc}{section}{\listtablename}
\listoftables

\cleardoublepage

\phantomsection
\addcontentsline{toc}{section}{\lstlistlistingname}
\lstlistoflistings

\phantomsection
\addcontentsline{toc}{section}{Nederlandstalige samenvatting}
\section*{Nederlandstalige samenvatting}
Een gekend probleem bij het schrijven van code is het vinden van lijnen code die niet bereikbaar zijn \cite{713323, 10.1145/1292316.1292319}. Onbereikbare code kan veel oorzaken hebben maar één van de moeilijker te analyseren situaties zijn condities die nooit voldaan worden. In deze gevallen is de code omringd door de conditie niet bereikbaar en zou deze dus eigenlijk weg moeten.

Niet bereikbare code is een probleem en kan zelfs kritiek worden in energie gelimiteerde omgevingen zoals ingebedde systemen \cite{embeddedefficiency}. In deze situaties is het vitaal dat onze code zo efficiënt mogelijk is en dus is er absolute nood aan een methode die niet bereikbare code kan detecteren.

Continue één counter automaten zijn hiervoor een zeer interesante oplossing, aangezien het bewezen is dat de bereikbaarheid aantoonbaar is voor deze automaton \cite{succintandparametriconecounter, danialandjoel, blondin2021continuous}. In dit onderzoek zullen we een praktische implementatie geven voor de theoretische werkwijze gegeven door Michael Blondin et al. \cite{blondin2021continuous}. Deze theoretische implementatie was specifiek opgezet om gebrek aan bereikbaarheid te kunnen aantonen.

De beperking om maar één parametrische klok (of counter) toe te staan heeft een groot effect op het bereik aan code dat we kunnen analyseren. Dit is echter onvermijdelijk aangezien er geen bewijs is voor de bereikbaarheid van continue automaten met twee counters. Voor drie counters en meer is er bewezen dat dit niet beslisbaar is \cite{FEARNLEY201526}.

In dit onderzoek zulen we een werkwijze geven om C code om te zetten naar continue één counter automaten. Gebruik makend van deze gegenereerde automaten zullen we analyseren of er in deze code lijnen zijn die niet kunnen bereikt worden. Dit zal gebeuren aan de hand van de volgende stappen:

\begin{enumerate}
	\item Voer pre analyse uit en analyseer alle compiler specifieke code
	\item Converteer de C code naar een parse tree met behulp van ANTLR
	\item Voer context vrije reducties uit op de parse tree
	\item Voer context sensitieve reducties uit op de parse tree
	\item Verifieer of de parse tree converteerbaar is naar een automaat
	\item Genereer de automaat
	\item Analyseer de bereikbaarheid
	\item Raporteer onbereikbare code indien deze gevonden is
\end{enumerate}

De code zal, buiten het feit dat deze maar één conditionele variabel kan hebben, ook beperkt worden in de operaties die hierop kunnen uitgevoerd worden. Aan een conditionele variabel kunnen enkel constante of parametrische waarden toegewezen worden of combinaties van sommen en aftrekkingen met constante of parametrische waarden. De conditionele waarde mag verder enkel met parameters of constante waarden vergeleken worden met de condities is gelijk aan, is niet gelijk aan, is strikt kleiner dan, is kleiner dan, is groter dan en is strikt groter dan. Ten slotte mogen de parameters zelf nooit aangepast worden tijdens de evaluatie van de functie.

Niet alleen de functie inhoud maar ook de definitie zal een beperking krijgen, zo accepteren we enkel parameters van het type integer.
\newpage
\section{Introduction}
A well known problem in writing code is determining whether or not every line of code is reachable. It is often impossible to properly determine whether or not a line of code guarded by a conditional statement is going to be reachable in any possible execution of the code \cite{713323, 10.1145/1292316.1292319}. Unreachable code is never desired and is an ever more pressing issue for energy restrained environments such as embedded computer systems \cite{embeddedefficiency}. In these systems, we want our execution to be as efficient as possible, and therefore avoid conditions that can never be satisfied or lines that can never be reached. To remedy this, a proper unreachable code detection approach is needed.

In order to analyse the reachability, we will make use of the fact that code is easily convertible to automata. All statements within the code that are related to counter updates can directly be converted to operations within an automaton. Each line of code can be represented by one or more nodes and a traversal of an automaton can be directly linked to the traversal of the code itself.

Continuous one counter automata are especially interesting as it has been proven that reachability is decidable for this type of automata \cite{succintandparametriconecounter, danialandjoel, blondin2021continuous}. In this paper we will provide a practical implementation for the theoretical approach that was provided by Michael Blondin et al.\cite{blondin2021continuous} which can be used to prove reachability of nodes.

Adding the restriction of only allowing a single parametric clock has a big effect on the types of code we can analyse. However, this is unavoidable, since there is no proof for continuous two counter machines. Three counter machines and up have been proven to be undecidable \cite{FEARNLEY201526}.

In this paper we will provide the overview of an approach which will convert C code to a continuous one counter automaton. Using the result of this first step, we will then provide a means to analyse the reachability of any of the nodes of said automaton. The following steps are taken without our approach to prove the existence of dead code.

\begin{enumerate}
	\item Perform pre analysis to resolve all C macro's
	\item Convert C code into a parse tree, using ANTLR
	\item Perform context free reductions on this parse tree
	\item Perform context sensitive reductions on this parse tree
	\item Validate the resulting parse tree, to be conforming to the code constraints
	\item Generate the counter automaton
	\item Analyse which nodes are reachable
	\item Link potential unreachable statements back to the lines within the original code base
\end{enumerate}

@@

Other than the described code constraints, functions must have integer parameter types (if any) to be converted into a counter automaton.

\subsection{Research questions}
The objective of this research can be summarised in two core research questions:

\begin{description}
	\item[Question 1: ] \hfill \\ Can we practically determine the reachability of continuous one counter automata?
	\item[Question 2: ] \hfill \\ Can we identify dead code in software written in the C language by analysing the reachability of the corresponding counter automata?
\end{description}

To solve these questions we will give two different approaches that are capable of determining reachability of nodes in both parametric and non-parametric one counter automata. Both approaches will additionally be applied to xrdp, a project used to login to remote machines that is entirely written in C.

The solution to the questions above can be divided into three different steps, where the final step will be further divided in two different solutions.\\
\\
\noindent
\textbf{Convert the C code to a one counter automaton} \\
This approach was previously done in a separate research but the general approach will be repeated here. It will furthermore be modified and perfected as an improvement of the existing work. This approach will be discussed in Section \textit{\ref{sec:generation} \nameref{sec:generation}} in which we will discuss the different steps part of converting C code.\\
\\
\noindent
\textbf{Specify a textual format for automata to allow reading and writing of automata}\\
The result of the previous step will be stored using the dot format which is part of Graphviz \cite{10.1007/3-540-45848-4_57}. This input should be readable to perform the reachability on it. We furthermore want to be able to provide this approach as a stand-alone concept, thus allowing the reachability analysis to be performed on arbitrary automata without the coupling with the compiler. In the Section \textit{\ref{sec:automaton input} \nameref{sec:automaton input}} we will discuss the steps needed to read an automaton from a dot file.\\
\\
\noindent
\textbf{Perform reachability analysis on non-parametric continuous one counter automata} \\
The first reachability analysis approach will be used to analyse the reachability of non-parametric continuous one counter automata. This implies that there can never be any operation applied to a counter with parameters involved. This approach will iteratively go through the automaton by making use of reachability intervals. The different steps and operations will all be discussed in detail in Section \textit{\ref{sec:reachability non parametric} \nameref{sec:reachability non parametric}}.\\
\\
\noindent
\textbf{Perform reachability analysis on parametric continuous one counter automata} \\
A secondary approach for reachability analysis will be given in Section \textit{\ref{sec:reachability parametric} \nameref{sec:reachability parametric}}. This approach will estimate the reachability intervals which were also used for the non-parametric variant. We will define conditions on the intervals and employ a SAT solver to guess a valid reachability interval configuration. The SAT solver will also support parametric automata and this approach will therefore be used as an alternative to the first approach in case the function uses parameters.

\section{Motivation}
Code efficiency is one of the core aspects on which the quality of code is measured. Inefficient code can incrementally lead to a slower product, even though it only represents a small part of a much larger whole. This is because code reuse is a critical aspect in Object Oriented Programming (OOP) \cite{10.1145/71605.71612}. In OOP we tend to write code by encapsulating it within classes (objects) based on the functionality they provide. This allows us to simply reuse that object whenever we need that specific functionality. As a direct result of this, every time we introduce code inefficiency, this will get replicated whenever we need use of the specific functionality in which the inefficiency was introduced.

Code inefficiency can also indirectly affect other projects. Many languages such as C, C++, Java and Python support the option to introduce libraries. These libraries are other projects which will entirely be introduced within the current project as to prevent the need to add a new implementation for a functionality which has already been implemented elsewhere. The same effect can occur here where an inefficiency within the library will get propagated to the other projects which make use of said library. As a means to speed up development there are even tools such as Hunter \cite{10.1145/2950290.2983934} which are intended to query large code bases to find code which already implements the feature that is needed, thus increasing the likelihood of using existing code.

One of the contributors to the efficiency of code is the reachability of code. Dead code implies that there is a piece of code that can never be reached and is thus useless. This might be due to a certain function simply never being invoked and in most cases this will not affect the actual efficiency of the code as this is trivially filtered out by compilers. However, guarded (conditional) pieces of code can be an issue in case they are not satisfiable. This can not simply be resolved by a compiler nor will it always get caught by all IDE's (Integrated Development Environment). A dead conditional statement will get evaluated every time said piece of code is used and can therefore contribute to a major efficiency reduction considering the code reuse that was mentioned above.

Popular IDE's do already implement reachability analysis up to a certain degree. However, the dead code elimination they perform is shallow and limited. The code example given below was executed using Clion and Eclipse, two of the more commonly used C IDE's. While it is known that both can find some dead code, it was not able to resolve this case, where the condition on line 4 can never evaluate to true. This is because the for statement on line 3 will always prevent line 4 from evaluating to true and yet they did not detect this.

\begin{lstlisting}[style=CStyle, caption={Example C code where Eclipse and Clion would not identify dead code.}]
bool is_mul(int a, int b) {
  int temp = b;
  for(; temp > 0;){
    if(temp == 0){
      break;
    }
    temp -= a;
  }
  return temp == 0;
}

@@

@@

In the past research has been done regarding dead code analysis. In 1994 there was a research done by Jens Knoop et al. \cite{10.1145/773473.178256} where they attempted to analyse partial dead code. Partial dead code is code that is only dead in some of its execution paths. It will aim to resolve this by changing the branching structure or the semantics of the program which will result in the bit of code no longer being dead in any program path. This resolves one of the cases that is not touched by the approach prescribed within this paper and they can therefore both provide benefits to projects without any redundancy in evaluation.

A different research has been done on the analysis of the data model behind software in an attempt to identify dead code at the declaration level. This was done in 1997 by Yih-Farn R. Chen et al. \cite{10.1145/267896.267924} and was applied on C++ projects. It concluded by showing that it was in fact capable of identifying a significant amount of dead functions and variables. This is however on a different level than the current research will work on. The research in question did analysis on the data model and might have therefore missed a significant amount of dead code that was introduced within the code itself.

As can be concluded from the examples given above, there has been research on dead code analysis. This research is however dated and it did not specifically cover the environment which we will cover within this research. It is furthermore shown that within the majorly used IDE's such as Eclipse and Clion the cases which we will attempt to detect with this new approach are not yet covered. We can conclude that this research brings additional value to what is known about dead code analysis and can further enhance the optimization of software.
\section{Preliminaries}
This paper will make use of counter automata and more specifically continuous one counter automata. These automata are derived from the general definition of automata and signify an extension by adding the notion of counters.

@@

At all times the automaton will be in a configuration $(q, c)$ where $q \in Q$ and $c$ the counter valuation where $(q_0, 0)$ is the initial configuration. The counter valuation $c \in \mathbb{N}$ represent one single value for which the counter is evaluated. We are interested in the reachability sets for each state q where the reachability set for q is $\{v | (q, v) \text{ is a reachable configuration}\}$.

The $\tau$ parameter represents a function that maps all nodes to their corresponding bounds. For a counter value to be part of the reachability interval of a node, it must respect the bounds implied by the $\tau$ function. The $\tau$ entry for a node q is represented by the an interval $[\underline{\tau(q)}, \overline{\tau(q)}]$ where $\underline{\tau(q)}$ and $\overline{\tau(q)}$ represent the lower and upper bounds of a node respectively.

The set of supported constraints within our evaluation are visualised by the $cond$ set below with $\epsilon$ representing the empty condition. Note that this is not binding for the code that we can analyse, several conditions might be convertible to constructs using only supported statements.
\[
cond = \{\leq c, =c, \geq c, \leq p, =p, \leq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

Parametric counter automata, as compared to regular counter automata, allow counters to be modified by parameters as well as constants. As the counters are assumed to be integers, we will only allow integer type parameters. We will consider the following set of allowed operations where $P$ is the set of parameters, the $\epsilon$ symbol will be used to represent an empty label.
\[
op = \{+c, -c, +p, -p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

The final state is omitted from this definition as we are not interested in this. In general the reachability of every line of code will be under scope and not just a single one. Furthermore, making the assumption that the final state is the only state of which we want to analyse reachability creates a significant reduction in the use cases of this approach, and was therefore discarded.

\subsection{COCA reachability}
In this section we give an overview of how we approach the concept of reachability. The provided concept was proven by Michael Blondin et al. \cite{blondin2021continuous} and will be the basis for the remainder of the approach.

Throughout the evaluation we will be tracking the reachability interval(s) of all nodes denoted by $reach(q)$. $Reach(q)$ will represent the set of possible counter values c that can exist in a configuration $(q, c)$.

A $run\;\rho$ is a sequence of configurations $(q_0, v_0)(q_1, v_1)...(q_n, v_n)$ such that the following holds with $T$ the set of all transitions:

\begin{itemize}
	\item $\forall i \in [0, n)\;\exists \alpha \in (0, 1]\;\exists t \in T: (q_i, v_i) \xrightarrow{\alpha\;t}(q_{i+1}, v_{i+1})$
	\item $\forall i \in [0, n]: v_i \in reach(q_i)$
\end{itemize}

A state $q_i$, and thereby represented line $l$, will be considered reachable if there exists a $run\;\rho$ with the sequence $(q_0, v_0)(q_1, v_1)...(q_i, v_i)$. We can therefore convert the reachability of lines of code to the the following problem. \\
\\
\noindent
\textbf{COCA reachability:} \\
Given a COCA $\mathcal{A}$, an initial configuration $(q_0, v_0)$ and a target configuration $(q_n, v_n)$ does there exist a run $\rho = (q_0, v_0) ... (q_n, v_n)$?\\
\\
In this paper we will give two different approaches that will approximate the $reach(q_i)$ intervals for each node $q_i \in Q$, thus attempting to prove whether or not such a run exists. This is equivalent as the following results from the definition given above:

\begin{itemize}
	\item $reach(q_i) \neq \emptyset \rightarrow \exists \rho = (q_0, v_0) ... (q_i, v_i)$
\end{itemize}

\section{Analysis of dead code in C}
The approach given within this paper consists out of two major phases. The first phase will consist of the conversion from C code to one counter automata. The second phase will use the result from the first phase, and perform a reachability analysis on these generated automata. The first phase was already covered in preceding research and the majority of the notes on the approach in the first phase were taken from that research. They were copied here as this research will conclude the unfinished work that was done there, as well as make continuous references to the flaws which arose from it.

\begin{figure}[b]
	\begin{tikzpicture}[auto, >=latex, font=\footnotesize\scshape]
	\tikzstyle{center} = [draw=black, rectangle, text width=4.6em]

	\node[center]	at (0, 0) 		(q0) 	{Resolve macros};
	\node[center]	at (3.2, 0)		(q1)	{Generate tree};
	\node[center]	at (6.4, 0)		(q2)	{Optimize tree};
	\node[center]	at (9.6, 0)	(q3)	{Validate conditions};
	\node[center]	at (8, -2)		(q4)	{Generate \\automaton};
	\node[center]	at (11.2, -2)		(q5)	{Discard \\tree};

@@

	\end{tikzpicture}
	\caption{Overview of the automaton conversion process}
\end{figure}

The first step of the conversion will be the preprocessing step in which all c compiler macro's will get resolved.

Next, the second step starts with an initialization, which consists of the generation of a parse tree. This will be done using the default C grammar, with the extension of the boolean type as this is not within the C language by default.

Using the resulting parse tree we will perform a simple context-free optimization where we will remove all needless nodes. This is done as an intermediate step as it will allow us to make assumptions in the second part of the optimization, where we will do a more in depth optimization cycle. This second part will perform optimizations based upon the context in which the nodes appear, and will try to simplify the syntax tree as much as possible. The resulting tree will contain no more potentially ambiguous nodes and all remaining nodes will be those relevant in the remainder of the evaluation.

The fourth step will be an analysis on whether or not the resulting abstract syntax tree conforms to all earlier specified requirements and, if this is the case, we will continue with the fifth step, which will be the generation of the counter automaton.

For a function to conform the arguments (if any) must be of type integer and must never be mutated within the function body, there can only be one counter and the counter can only be mutated using the operations part of the $op$ set displayed below.
\[
op = \{+c, -c, +p, -p, \leq c, =c, \geq c, \leq p, =p, \geq p: c \in \mathbb{N}, p \in P\} \cup \{\epsilon\}
\]

@@

\subsection{Preprocessing}
\label{sec:conversion}
The program will start by resolving all the compiler macros that are part of C. These are intended to be resolved by the compiler at compile time, and are often dependent on which compiler was chosen/the environment on which the code is executed. For our use case I decided to select the global variables in such a way that these represent a windows machine that is using 64 bits. This is however not at all restricted to that and could further be expanded to support any environment and any combination of global variables possible. This selection is in essence not relevant as it should not affect the presence of dead code.

Something else that is part of these macros are the include statements. These statements were partially resolved in the sense that only the imports from the current project were handled. Includes from global C libraries are not relevant for the type of dead code elimination we attempt to do within this project and we therefore decided that the addition of said libraries would only result in a much larger code base to be analysed without any additional benefits for the dead code analysis of the actual project.

The if/elif/else macros were simply resolved by converting them to boolean expressions which can then be resolved by any boolean solver. These macro's are intended to identify whether or not global variables are defined and/or are of a current value and are thus relatively simple to resolve.

Finally, the define and undefine macros were resolved by simply tracking the values they defined and/or undefined throughout the evaluation of the code.

\subsection{Initialization}
Using the flattened result of the preprocessor, the next step will start by running the generator ANTLR compilation unit on the given C code. In case there are any issues regarding simple compilation rules, the program will exit with an error, indicating the issues in the code. For the remainder of the execution, the code is assumed to be correct, and there will be no regard for potential issues within the code itself.

@@

However, the creation of nodes is not the same in all situations, in some cases, a specific kind of node will be created and in other cases it will not, depending on the rule used for the current node (which we can determine from the context of the current node). In order to know whether or not we need to pop a node from the node stack, we will allow the code to look at the top node in the stack, and evaluate the type of set node. No other context related evaluations will be allowed.\\
\\
\noindent
\textbf{Context-free analysis assignmentExpression example}\\
To look at a concrete example, consider the following rule in the grammar:
\begin{align*}
	&assignmentExpression\\
	&\text{:}\indent \text{conditionalExpression}\\
	&|\indent \text{unaryExpression assignmentOperator assignmentExpression}\\
	&|\indent \text{Digitsequence}; \\
\end{align*}
The rule for conditionalExpression needs to be there, as there is a chain of expressions, and some expressions need to be higher in the order than others, therefore all expressions with a higher priority need to be evaluated before the lower priority expressions. This is done by adding a rule to a lower priority expression in each expression class. It is however undesirable for these nodes to be added. Without regard for context in other nodes, we can simply look at the current node to see whether or not the conditionalExpression line is used (which ANTLR generated context allows us to do).

@@

Important to note is that no folding/substitution will occur within loops, as these operations occur a variable number of times, and since there is no way of determining the exact number of times without evaluating the expression, this is considered too complex and often impossible, as counters may depend on parameters, which need to be chosen at evaluation time.\\
\\
\noindent
\textbf{Context-sensitive reduction example: constant propagation}\\
Another example of a needed reduction, is constant \textit{propagation} and \textit{folding}. It is possible that a code segment would initially be rejected, due to use of unwanted variables, while in essence, these variables are nothing but variable representations of results of constant expressions. By \textit{folding} we will attempt to simplify constant expressions by replacing them with their results. Once simplified to a constant value we will then attempt to \textit{substitute} the variables by replacing further uses with said value.

Consider the code in Listing \ref{lst:divide} below. It would initially be rejected, because the \textit{counter} variable is initialized with a value that is neither a constant nor a parameter.

\begin{lstlisting}[style=CStyle, caption={Example of a function in need of constant propagation}, label={lst:divide}]
bool divide(int p, int n){
	int variable = 5;
	variable += 20;
	int counter = variable;
	while(counter > 0){
		counter -= n;
	}
	counter += 1;
	counter -= 1;
	if(counter == 0){
		return true;
	}
	return false;
}
\end{lstlisting}

However, due to constant propagation and folding, we can simplify the two statements on line 2 and line 3, so that they become one simplified operation. This can be seen in Listing \ref{lst:dif variable}.

\begin{lstlisting}[style=CStyle, caption={Result of simplifying operations on \textit{variable}}, label={lst:dif variable}]
int variable = 25;
\end{lstlisting}

Now that \textit{variable} is just a constant, we can substitute its value which leads us to the situation given in Listing \ref{lst:sub variable}.
\begin{lstlisting}[style=CStyle, caption={Result of substituting \textit{variable}}, label={lst:sub variable}]
int variable = 25;
int counter = 25;
\end{lstlisting}

Now \textit{variable} has no further benefit for the execution and will be dropped.

\begin{lstlisting}[style=CStyle, caption={Final version of the simplified function}, label={lst:divide final}]
bool divide(int p, int n){
	int counter = 25;
	while(counter > 0){
		counter -= n;
	}
	counter += 1;
	counter -= 1;
	if(counter == 0){
		return true;
	}
	return false;
}
\end{lstlisting}
